Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-07-20 manuscript.Rnw
more writing

The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}. Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies\citep{Falster2015}, we here show that partitioning between stems and leaves varies strongly between PFTs (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm). We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}. In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.

@@

The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\citep{ise2010}. Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}. Globally, forests store approximately 360Pg of carbon in living biomass\citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}. Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014}.

@@

%paragraph on LMF, it is an interesting variable because it is an important component of turnover of biomass. Link to leaf lifespan across PFTs as well. 
For the above-ground biomass pools, we calculated the leaf mass fraction ($M_F / M_T$, where $M_T$ is total above-ground biomass) and leaf area ratio ($A_F / M_T$). These variables are related by,

@@

For LMF we hypothesize that it is higher for EG than DA, with EA taking an intermediate value as determined by its mean LMA. This would be consistent both with leaf lifespan correlate as well as minimizing variation in leaf area per unit biomass (which determines light interception).

We also wonder whether differences in LMF and LAR (and its component variables) are sensitive to climate. One may expect this to be the case because short-term experiments have shown responsiveness of biomass partitioning to soil water etc. etc.; or this variation may wash out at a large scale. It is thus the question whether variation within PFT can be explained by climate or not. Indeed, many GVMs assume that climate affects biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Methods \ref{sec:gvmsummary}).

Here, we use a unique, new database\citep{Falster2015} (Fig. 1) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms. The dataset includes over 21,000 observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig. S1).

One key challenge is to account for the very large size variation in any allometric variable, also LMF. We do this by fitting an assumption-free model (GAM) and studying PFT patterns at a common height.

@@

\subsection*{Terminology}
\label{sec:partterm}

The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \citep{litton2007}. Here, we define biomass partitioning as the actual distribution of biomass between compartments (e.g., leaves vs. stems), and biomass allocation as the proportion of net primary production (NPP) that is allocated to some compartment. The two processes are different because of continuous turnover of biomass, which differs strongly between compartments. We may write \citep{McMurtrieWolf1983},

@@

Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}. We then found that LMF at a common height was proportional to the leaf construction cost, as measured by leaf mass per area (LMA) (Fig. 2b). As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig. 2c and S3).

Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$). This ratio is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}.  Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig. 2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig. 2e).

@@

paragraph - variation in LMF was the same as MF/AS, suggesting that MT/AS is a fairly conserved quantity \ref{eq:M2}. Consistent with \citep{chave2005} who showed that aboveground biomass can be well estimated with Ht*AS*wood density. This works because, for large plants, most of aboveground biomass is wood.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure1.pdf}
    \caption{\textbf{Overview of the global biomass and allometry database.} (a) Variables were measured on \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species. (b) Coverage of the dataset across global climate space. Grey hexagons indicate the number of 0.5\degree~cells with woody vegetation across the space. Colour symbols show the locations of sampled individuals for three dominant woody functional types.}
    \label{fig:figure1}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
    \caption{\textbf{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost. } (a) Leaf mass fraction ($M_F/M_T$ = leaf mass / above-ground biomass) by PFT. Each symbol is an individual plant. Lines are generalized additive model fits. (b) and (c) Leaf mass fraction and leaf area ratio ($A_F/M_T$) at the average plant height in the dataset, estimated from panel (a). (d) Average leaf mass per unit basal stem area, and (e) leaf area per unit basal stem area for the three major PFTs confirm that the between-PFT variation in leaf mass fraction is due to leaf mass per unit basal area. Error bars are 95\% confidence intervals. Letters denote significant differences (at $\alpha = 0.05$).}
    \label{fig:figure2}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \caption{\textbf{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning}. a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.  DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms. Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect. Letters denote significant differences (at $\alpha = 0.05$). b-c) Relationships between climate and $M_F/A_S$ within each PFT. Solid lines are generalized additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
    \label{fig:figure3}
\end{figure}

@@

Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types. However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL). Typically, parameters related to biomass allocation differ between PFTs (often based on little empirical evidence), which then lead to differences in biomass partitioning. We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them. We base this classification on a recent model intercomparison study \citep{dekauwe2014}. Note that we use 'partioning' to indicate the actual distribution of biomass among plant components, and 'allocation' the fraction of net primary production to each component (see Methods \ref{sec:partterm}). Finally, we briefly summarize how current GVMs include effects of climate on biomass partitioning.

@@

Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots). CABLE, EALCO \citep{dekauwe2014} and IBIS \citep{foley1996} are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep  (though, for both models, alternative model formulations have been used as well). Constant allocation fractions still give rise to differences in biomass partitioning between PFTs (see Methods \ref{sec:partterm}), since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.

>>>>

The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs). Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies, we here show that partitioning between stems and leaves varies strongly between PFTs (as defined by deciduous vs. evergreen and gymnosperm vs. angiosperm). We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum. In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.

@@

The partitioning of forest biomass among leaves and stems strongly influences the productivity and carbon cycle of the world's vegetation\citep{ise2010}. It affects the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}. Biomass stored in woody stems has a long residence time \citep{luyssaert2008}, whereas leaf biomass turns over quickly, entering the soil carbon cycle where the majority of carbon is released back to the atmosphere \citep{ryanlaw2005}. Globally, forests store approximately 360Pg of carbon in living biomass\citep{panetal2011}, equivalent to almost 40 years of current anthropogenic CO\textsubscript{2} emissions \citep{friedlingsteinetal2014}. Reducing uncertainties about biomass partitioning and carbon residence times in GVMs is therefore a priority for understanding the effects of climate and other environmental change on the global carbon cycle \citep{friend2014,dekauwe2014}.

@@

Despite considerable advances in allometric theory (enquist etc. etc.) we don't have a clear picture of differences between PFTs. Instead, most of the work has focussed on understanding size-related shifts in biomass partitioning.  Longer leaf lifespan would logically lead to higher LMF (as argued by enquist, niklas) only if the annual allocation does not vary with leaf lifespan, but there is no a priori reason why this should be so. We take the view that a broader evidence base is needed to elucidate patterns which can subsequently be used to refine and improve existing models of allometry.

In this study we focus on the partitioning of above-ground biomass between short-lived leaves and long-lives woody pools. We are also interested in a related variable describing the amount of leaf area per unit aboveground biomass, which is more closely related to light interception and hence productivity. These variables are related as follows,

@@

Where $M_T$ is total above-ground biomass, $M_F$ total leaf biomass, $A_F$ total leaf area, the leaf mass fraction is $M_F / M_T$, and leaf area ratio is $A_F / M_T$.  For LMF we hypothesize that it is higher for EG than DA, with EA taking an intermediate value as determined by its mean LMA. This would be consistent both with leaf lifespan correlate as well as minimizing variation in leaf area per unit biomass (which determines light interception).

We also wonder whether differences in LMF and LAR (and its component variables) are sensitive to climate. One may expect this to be the case because short-term experiments have shown responsiveness of biomass partitioning to soil water etc. etc. Alternatively observed variation in LMF across the globe may be more closely related to PFT. Indeed, many GVMs assume that climate affects biomass partitioning within a given PFT, often through soil water stress or other abiotic stress factors (see Methods \ref{sec:gvmsummary}). Approaches taken by these models differ widely, thus it is pertinent that global-scale patterns are described to decide between the different assumptions.

Here, we use a unique, new database\citep{Falster2015} (Fig. 1) to establish general rules on how biomass partitioning differs among three dominant woody PFTs across the globe: evergreen gymnosperms, evergreen angiosperms, and deciduous angiosperms. The dataset includes over 21,000 observations of biomass and size metrics for individual plants, contributed by 175 studies across nine vegetation types (Fig. 1). Since we are here interested in the differences between PFTs and relation to climate, we do not study the dependence on size per se. We do this by fitting an assumption-free model (GAM) and studying PFT patterns at a common height.

@@

\subsection*{Terminology}

The terms 'partitioning' and 'allocation' have been used in various ways, confusing comparisons between studies \citep{litton2007}. Here, we define biomass \emph{partitioning} as the actual distribution of biomass between compartments (e.g., leaves vs. stems), and biomass \emph{allocation} as the proportion of net primary production (NPP) that is allocated to some compartment. The two processes are different because of continuous turnover of biomass, which differs strongly between compartments. We may write \citep{McMurtrieWolf1983},

@@

Because of allometric relationships between the studied variables \citep{niklas1994}, there was a strong effect of plant height on LMF (Fig. 2a and S2). We corrected for plant height by estimating LMF at a common plant height, using a generalized additive model, a semi-parametric method that makes no assumption on the underlying functional relationships \citep{wood2006}. We then found that LMF at a common height was proportional to the average leaf mass per area (LMA) across the three PFTs (Fig. 2b). As a consequence, LAR was invariant between PFTs, because LAR = LMF / LMA (Fig. 2c and S3). Mirroring the results for LMF, we found that the amount of leaf mass per unit stem area ($M_F/A_S$) differs among the three PFTs, that these differences are also proportional to LMA (Fig. 2d), and that PFTs are similarly invariant in the amount of leaf area per unit stem area ($A_F/A_S$) (Fig. 2e).

@@

paragraph - variation in LMF was the same as MF/AS, suggesting that MT/AS is a fairly conserved quantity \ref{eq:M2}. Consistent with \citep{chave2005} who showed that aboveground biomass can be well estimated with Ht*AS*wood density. This works because, for large plants, most of aboveground biomass is wood. Another key metric governing vegetation behavior is the amount of leaf maintained per unit cross-sectional area of stem ($A_S$). This ratio is relevant because plant leaf mass is frequently estimated from $A_S$ with allometric equations \citep{shinozaki1964b, chave2005}.

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure1.pdf}
    \caption{Overview of the global biomass and allometry database. (a) Variables were measured on \Sexpr{nrow(dataset)} individual plants from \Sexpr{length(unique(dataset$species))} species. (b) Coverage of the dataset across global climate space. Grey hexagons indicate the number of 0.5\degree~cells with woody vegetation across the space. Colour symbols show the locations of sampled individuals for three dominant woody functional types.}
    \label{fig:figure1}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure2.pdf}
    \caption{Dominant woody PFTs differ in leaf mass fraction due to underlying differences in leaf construction cost. (a) Leaf mass fraction ($M_F/M_T$ = leaf mass / above-ground biomass) by PFT. Each symbol is an individual plant. Lines are generalized additive model fits. (b) and (c) Leaf mass fraction and leaf area ratio ($A_F/M_T$) at the average plant height in the dataset, estimated from panel (a). (d) Average leaf mass per unit basal stem area, and (e) leaf area per unit basal stem area for the three major PFTs confirm that the between-PFT variation in leaf mass fraction is due to leaf mass per unit basal area. Error bars are 95\% confidence intervals. Letters denote significant differences (at $\alpha = 0.05$).}
    \label{fig:figure2}
\end{figure}

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.99\textwidth]{Figure3.pdf}
    \caption{Beyond influencing the distribution of PFTs, climate only modestly influences biomass partitioning. a) Leaf mass per unit stem cross-sectional area ($M_F/A_S$) for each PFT within three different biomes.  DA: deciduous angiosperms, EA: evergreen angiosperms, EG: evergreen gymnosperms. Error bars are 95\% confidence intervals for the mean, estimated from a linear mixed-effects model using species within study as the random effect. Letters denote significant differences (at $\alpha = 0.05$). b-c) Relationships between climate and $M_F/A_S$ within each PFT. Solid lines are generalized additive model (GAM) fits (with a base dimension of 3); with grey areas indicating the 95\% confidence interval around the GAM.}
    \label{fig:figure3}
\end{figure}

@@

Most leading GVMs assume that biomass partitioning between leaves and woody stems differs between major plant functional types. However, there is no consensus on how these differences are parameterized, or whether they are linked to inherent differences in leaf mass per area (LMA) or leaf lifespan (LL). Typically, parameters related to biomass allocation differ between PFTs (often based on little empirical evidence), which then lead to differences in biomass partitioning. We briefly summarize three classes of biomass allocation models, and how differences in biomass partitioning between PFTs arises in each of them. We base this classification on a recent model intercomparison study \citep{dekauwe2014}. Note that we use 'partioning' to indicate the actual distribution of biomass among plant components, and 'allocation' the fraction of net primary production to each component (see Methods). Finally, we briefly summarize how current GVMs include effects of climate on biomass partitioning.

@@

Some models assume that a constant fraction of net primary production (NPP) is allocated to each component (leaves, stems, roots). CABLE, EALCO \citep{dekauwe2014} and IBIS \citep{foley1996} are examples of models that allocate a PFT-specific fraction of NPP to each plant component each timestep  (though, for both models, alternative model formulations have been used as well). Constant allocation fractions still give rise to differences in biomass partitioning between PFTs (see Methods in main text), since turnover of leaves (determined by LL) will always differ between at least deciduous and evergreen PFTs.
